---
title: "Biostatistics vignette"
author: "Rob Knell"
date: "January 2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Biostatistics vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(Biostatistics)
```

# The Biostatistics Package

This package consists of a series of learnr tutorials for use in teaching statistics to biologists. They were written for use in undergraduate and postgraduate teaching in the UK but they could also be used for individual, self-directed learning. The subjects covered range from basic data visualistion and description through to reasonably advanced linear modelling. There are obviously many subjects which are not currently covered such as generalised linear models, mixed effects models and multivariate statistics and it is hoped that these will be incorporated in the future.

There is a strong emphasis throughout the tutorial on analysing real data sets. This is much better for learning statistics than using synthetic example data because with real data comes all of the issues and uncertainty associated with real science. The data used here have mostly been made publicly available by the authors of papers published in the biological literature, mostly via the [Dryad data repository](https://datadryad.org/stash), and I would like to thank all of them for this.

The tutorials are written for the [learnr](https://rstudio.github.io/learnr/) package which uses an [rmarkdown](https://rmarkdown.rstudio.com/) framework to render tutorials into [shiny](https://shiny.rstudio.com/) webapps. The rmarkdown files for all of the tutorials are available on the [author's github page](https://github.com/rjknell).

## Running tutorials

There are two ways of running these tutorials. The easy way assumes you are using a recent version of RStudio. If this is the case then once you have installed the package the tutorials will show up in the 'Tutorial' tab in the RStudio pane that also includes the Environment and History tabs. Click the "Start Tutorial" button and the tutorial will render, which can take a few seconds, and then appear in the Tutorial tab. You'll probably want to maximise the pane within your RStudio window. If you want to finish the tutorial click on the 'Stop' sign button at the top left of the tab.

If you would rather run your tutorial in a separate browser window then you can use the `run_tutorial()` function from the learnr package. You need to specify the name of the tutorial and the package, so 

`learnr::run_tutorial("02 Descriptive statistics", package = "Biostatistics")`

will run the Descriptive Statistics tutorial and 

`learnr::run_tutorial("17 Linear models 5 Multiple Regression", package = "Biostatistics")`

will run the Multiple Regression tutorial. In my experience the first method, with the Tutorial pane, seems more stable and sometimes tutorials won't render using `run_tutorial()` for reasons that are not clear.

## List of tutorials

The tutorials currently in the package are:

00 Introduction                                             
01 Frequency histograms                                     
02 Descriptive statistics                                   
03 Boxplots                                                 
04 Scatterplots                                             
05 Sampling_distributions                                   
06 Standard_errors                                          
07 Confidence intervals                                     
08 Confidence intervals for the difference between two means
09 Paired-sample t-tests                                    
10 Two sample t-tests                                       
11 Chi-square tests                                         
12 Correlation                                              
13 Linear models 1 Single factor ANOVA                      
14 Linear models 2 Linear Regression                        
15 Linear models 3 Model assumptions and diagnostics        
16 Linear models 4 Multi factor ANOVA                       
17 Linear models 5 Multiple regression                      
18 Linear models 6 Factors and continuous variables         
19 Linear models 7 Model selection

## Datasets used in this package