---
title: "Descriptive and Exploratory Statistics: Introduction"
output: 
  learnr::tutorial:
    theme: default
    css: http://research.sbcs.qmul.ac.uk/r.knell/learnr_data/test2.css
runtime: shiny_prerendered
author: Rob Knell
description: >
  Introduction to the biostatistics tutorials.
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction to the biostatistics tutorials

Welcome to the biostatistics tutorials. These are a set of interactive tutorials developed for teaching statistics to biologists at both undergraduate and postgraduate levels. The tutorials cover a range of subjects from basic data visualisation and descriptive statistics through the fundamentals of statistical testing to advanced linear model fitting. There is a strong emphasis throughout on analysing real datasets in order to give experience with the messiness and uncertainty that can arise when deling with real data.

The concepts covered in these tutorials here are explained with both text and, in many cases, embedded video. There are interactive code exercises which are written for people with only a basic understanding of R, and there are quizzes that you can use to test your knowledge and understanding.

It is hoped in future to extend these tutorials to cover further subjects such as multivariate statistics, mixed effects models and generalised linear models.

Please note that the current release of RStudio will include some tutorials by defualt in the tutorials pane which are not a part of this package. These are: Data basics, Filter observations, Create new variables, Set Up and Hello, Tutorial! There is not currently an easy way to remove these.

These tutorials were written by Rob Knell, School of Biological and Chemical Sciences, Queen Mary University of London. <a href="mailto:R.Knell@qmul.ac.uk">Send Email</a>

## License

This content is licensed under a [GPL-3](https://www.gnu.org/licenses/gpl-3.0.en.html) license

## Interactive coding exercises

*Here's an example of a simple coding exercise with an empty box provided for entering the answer. Note that there is a "hints" button. Most of the exercises will have either this or a "solution" button. These are here to help you if you get stuck. If there is more than one hint then there will be a "next hint" button visible when you bring the first hint up. When you want to run the code click the "Run code" button on the right.*

Write the R code required to add two plus five:

```{r two-plus-two, exercise=TRUE}

```

```{r two-plus-two-hint-1}
# You can just use the number keys and the
# "+" symbol
```

```{r two-plus-two-hint-2}
# This is the solution:

2 + 5
```

*Some exercises come with code already in the box which you will need to edit.*

This code will draw a scatterplot of two sets of random numbers. Change the plot symbol colour from "darkgreen" to "steelblue".

```{r plot1, exercise=TRUE, exercise.lines = 5}
plot(rnorm(20) ~ runif(20),
		 pch = 16,
		 col = "darkgreen")
```

```{r plot1-hint-1}
# The argument that sets the colour is "col = "
# You just need to replace "darkgreen" with "steelblue"
# Don't forget the quote marks
```

```{r plot1-hint-2}
# This is the solution:

plot(rnorm(20) ~ runif(20),
		 pch = 16,
		 col = "steelblue")
```

## Multiple choice quizzes

These are largely self explanatory. Some quizzes are4 "single best answer", meaning that only one answer per set is correct. Some have more than one answer that is correct and if this is the case you'll be told it in the text.

## Tutorial tips

If you're running a tutorial in the tutorial pane of RStudio don't forget about the three buttons in the top left. The home button takes you back to the list of tutorials, the stop button ends the tutorial and the button in the middle labelled "Show in new window" opens the tutorial in a pop-up window. If you are using a small screen then the pop-up window can be a really good option.

Once a tutorial is finished it should be self evident, but if in doubt when you come to the license information then you've got to the end. The tutorials don't automatically follow each other so you need to close the one you've finished and then if you wish open the next one.

## 
